---
title: "Microbial Bile Acid Analysis - in vitro BA incubation in 37 sFMT strains"
author: "Min Soo Kim"
date: '`r format(Sys.time(), "%Y-%m-%d %H:%M")`'
output: 
  html_document:
    code_folding: show
    theme: spacelab
    number_sections: true
    highlight: monochrome
    fig_width: 11
    fig_height: 8.5
    toc: true
    toc_float: true
---
# Experiment description

This experiment measured bile acids in media containing 37 strains in Susan's sFMT community that was incubated with the following bile acids:CA-d5, DCA, CDCA-d5, LCA, GCA, TCA. The metabolomics experiment was run on ThermoFisher TSQ and processed on the Quan Browser in XCalibur


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up environment
```{r, echo=FALSE, message=FALSE}
library(data.table)
library(tidyverse)
library(readxl)
library(qiime2R)
library(ggbreak)
library(vegan)
library(ape)
library(gplots)
library(ggplot2)
library(ggtree)
library(ggnewscale)
library(stats)
library(patchwork)
#library(ggforce) #needs dependencies. ask jordan to download
theme_set(theme_q2r())

setwd("/home/msk5737/BileAcids/221211_BA_sFMT_invitro/")

```

# Read in raw data make data tables

## Read in the raw data and append sample and BA metadata
```{r, message=FALSE, warning=FALSE}

#read in raw data
Input<-readxl::excel_sheets("220117_BA_sFMT_invitro_QUAN_Short.XLS")
Input<-Input[!Input %in% c("Component","mdlCalcs")]

Rawdata<-
lapply(Input, function(i){
  read_excel("220117_BA_sFMT_invitro_QUAN_Short.XLS", sheet=i, skip=4) %>%
    dplyr::select(Filename, Area) %>%
    mutate(FeatureID=i)
    })


Rawdata<-Rawdata %>% do.call(bind_rows, .)

Rawdata<-
Rawdata %>%
  filter(!is.na(Filename) & !grepl("^Created|^User|^Quantis", Filename)) 

#read in BA metadata
BAinfo <- read_excel("BAmetadata.xlsx", sheet = 1) 

metadata<-
  BAinfo %>% 
  dplyr::select(FeatureID=QuanID, Index, Cal, IS, Detected)

metadata<-
metadata %>%
  left_join(metadata %>% dplyr::select(Cal=Index, Calibrator=FeatureID)) %>%
  left_join(metadata %>% dplyr::select(IS=Index, InternalStandard=FeatureID))


#clean up the raw data and add meaningful column titles
Rawdata<-Rawdata %>% mutate(Area=as.numeric(Area))

data<-
Rawdata %>%
  left_join(metadata %>% dplyr::select(FeatureID, InternalStandard, Detected)) %>%
  left_join(
  Rawdata %>%
  filter(FeatureID %in% metadata$InternalStandard) %>%
  dplyr::rename(InternalStandard=FeatureID, IS_Area=Area)
  ) 

data<-
data %>%
  mutate(Relative_Ratio=Area/IS_Area)

#take out undetected/unquantitated compounds
data <-
  data%>% 
  filter(Detected == "Y") %>% 
  dplyr::select(!Detected)

```

## Make cal curves and quantitate BAs
```{r, message=FALSE, warning=FALSE}
#make a table of cals and their concentrations
CalConc<-tibble(Filename=paste0("cal",1:10), StdConc=c(0.01, 0.025, 0.05, 0.1, 0.25, 0.5, 1, 2.5, 5, 10))

#pull out the cals from data
stds<-
  data %>%
  filter(grepl("^cal", Filename)) %>%
  left_join(CalConc)

#plot out the cals 
stds %>%
  ggplot(aes(x=Relative_Ratio, y=StdConc)) +
  geom_point() +
  facet_wrap(~FeatureID, scales="free")

#fit a linear curve on the plotted cals
stds %>%
  filter(FeatureID %in% metadata$Calibrator) %>%
  ggplot(aes(x=Relative_Ratio, y=StdConc)) +
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~FeatureID, scales="free")

#pull slopes to use for quantatitation
#note, come back and fix to not force through zero, or use weighted?
CalCurves<-
stds %>%
  filter(FeatureID %in% metadata$Calibrator) %>%
  group_by(FeatureID) %>%
  summarize(
    N_concentrations=n(),
    Slope=lm(Relative_Ratio~0+StdConc)$coefficients,
    R2=lm(Relative_Ratio~0+StdConc) %>% broom::glance() %>% pull(r.squared)
  )

#quantitiate BAs using the calculated slopes of the cal curves
data<-
data %>% 
  left_join(metadata %>% dplyr::select(FeatureID, Calibrator)) %>%
    left_join(CalCurves %>% dplyr::select(Calibrator=FeatureID, Slope)) %>%
  mutate(Concentration_uM=Relative_Ratio*Slope) %>%
  mutate(Concentration_zerofill=if_else(is.na(Concentration_uM), 0, Concentration_uM))

data<-
data %>%
  dplyr::select(Filename, FeatureID, Concentration_uM=Concentration_zerofill)

#merge in sample metadata
mergeddata<-
data %>%
  left_join(
    read_excel("220111_BA_sFMT_invitro_samplemetadata.xlsx") %>% 
      dplyr::select(Filename=SampleID, Group, ReplicateNum) %>%
      mutate(Filename=as.character(Filename))
  ) %>%
  filter(!grepl("^cal", Filename)) %>% 
  filter(!grepl("blank", Filename))

mergeddata<-
  mergeddata %>%
  filter(!grepl("blank", Filename))


mergeddata<-
  mergeddata %>%
  mutate(SampleID=paste(Group,ReplicateNum, sep="_"))
```

## Clean up the dataset to only show concentration and log2conc
```{r, message=FALSE, warning=FALSE}

#combine coeluters
mergeddata <- 
  mergeddata %>%
  mutate(FeatureID=gsub("TCA","TCA/THCA", FeatureID, fixed = TRUE)) %>%
  mutate(FeatureID=gsub("TUDCA","TUDCA/THDCA", FeatureID, fixed = TRUE)) %>%
  mutate(FeatureID=gsub("TDCA","TDCA/TCDCA", FeatureID, fixed = TRUE)) %>%  
  mutate(FeatureID=gsub("GUDCA","GUDCA/GHDCA", FeatureID, fixed = TRUE)) %>%  
  #mutate(FeatureID=gsub("UDCA","UDCA/HDCA", FeatureID, fixed = TRUE)) %>%  
  mutate(FeatureID=gsub("12-oxoDCA","12-oxoDCA/7-oxoCDCA/6-oxoHDCA", FeatureID, fixed = TRUE))

#make a data matrix by pivot_wider
BA_wide<-
  mergeddata %>%
  #filter(!grepl("Group", Group)) %>%  #take out when i have subgroups data
  dplyr::select(FeatureID, SampleID, Concentration_uM) %>%
  pivot_wider(names_from = SampleID, values_from = Concentration_uM) %>%
  as.data.frame() %>%
  column_to_rownames("FeatureID")

#QC Needed!


BA_wide<-BA_wide[rowSums(BA_wide)>0,]

BA_wide<-
  BA_wide %>%
  rownames_to_column("FeatureID") %>%
  pivot_longer(!FeatureID, names_to = "SampleID", values_to="Concentration_uM") %>%
  left_join(
    BA_wide%>%
      rownames_to_column("FeatureID") %>%
      pivot_longer(!FeatureID, names_to = "SampleID", values_to="Concentration_uM") %>%
      group_by(FeatureID) %>%
      filter(Concentration_uM==min_nonzero(Concentration_uM)) %>%
      ungroup() %>%
      dplyr::select(FeatureID, MinConcentration=Concentration_uM) %>%
      distinct()
  ) %>%
  mutate(log2Conc=log2(Concentration_uM + (2/3*(MinConcentration)))) %>%
  dplyr::select(FeatureID, SampleID, log2Conc) %>%
  pivot_wider(names_from = SampleID, values_from = log2Conc)

#make a data table with pivot_longer
BA_long <- 
  BA_wide %>% 
  pivot_longer(!FeatureID, names_to = "SampleID", values_to="log2Conc") %>% 
  left_join(mergeddata) %>% 
  dplyr::select(SampleID, Group, ReplicateNum, FeatureID, Concentration_uM, log2Conc)


BA_long %>% pivot_wider(c(FeatureID), names_from = SampleID, values_from = log2Conc)

BAorder<-
  BA_long %>% 
  left_join(BAinfo %>% dplyr::select(FeatureID=QuanID, BAorder)) %>% 
  group_by(FeatureID) %>% 
  summarize(mean=mean(BAorder)) %>% 
  arrange(mean) %>% 
  pull(FeatureID)

BAorderRev <-
  mergeddata %>% 
  left_join(BAinfo %>% dplyr::select(FeatureID=QuanID, BAorder)) %>%   
  group_by(FeatureID) %>% 
  summarize(mean=mean(BAorder)) %>% 
  arrange(desc(mean)) %>% 
  pull(FeatureID)
    

write_csv(BA_long, "230127_BA_sFMT_invitro_ConcData.csv")

#save an extra file of just JEB00356 for jingcheng
#BA_long %>% 
#  filter(Group %in% c("JEB00356", "Sterile")) %>% 
#  write_csv(., "JEB00356_BileAcids.csv")


```

## What are the most abundant bile acids?
```{r message=FALSE, warning=FALSE}
BA_long %>%
  group_by(FeatureID) %>% 
  filter(!is.na(Concentration_uM)) %>% 
  summarise(mean=mean(Concentration_uM),median=median(Concentration_uM),min=min(Concentration_uM),max=max(Concentration_uM),sd=sd(Concentration_uM),n=n()) %>% 
  arrange(desc(mean)) %>%
  interactive_table()

BA_long %>%
  arrange(Group, FeatureID) %>%
  group_by(FeatureID) %>% 
  filter(!is.na(Concentration_uM)) %>% 
  summarise(mean=mean(Concentration_uM),median=median(Concentration_uM),min=min(Concentration_uM),max=max(Concentration_uM),sd=sd(Concentration_uM),n=n()) %>% 
  arrange(desc(mean)) %>%
  ggplot(aes(x=FeatureID,y=mean))+
  geom_bar(stat="identity", color="black")+
  theme_q2r()+
  theme(axis.text.x = element_text(angle = 45,hjust = 1))
```

# Make some exploratory figures - PCA, biplot, indiv conc plots

##biplot and skim data
```{r, eval=FALSE}
pc<-
BA_wide %>%
  column_to_rownames("FeatureID") %>%
  t() %>%
  prcomp()

biplot(pc)


pc$x %>%
  as.data.frame() %>%
  rownames_to_column("StrainID") %>%
  separate(StrainID, c("StrainID", "Replicate"), sep="_") %>%
  ggplot(aes(x=PC1, y=PC2, color=StrainID)) +
  geom_point() +
  geom_line(aes(group=StrainID))



BA_wide %>%
  pivot_longer(!FeatureID) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(FeatureID=="TCA") %>%
  ggplot(aes(x=StrainID, y=value)) +
  geom_point() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~FeatureID, scales="free")



BA_wide %>%
  pivot_longer(!FeatureID) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(FeatureID=="TDCA") %>%
  ggplot(aes(x=StrainID, y=value)) +
  geom_point() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~FeatureID, scales="free")

pdf("skim.pdf", height=8, width=12, useDingbats=F)
for (analyte in BA_wide$FeatureID){
 p<-
   BA_wide %>%
  pivot_longer(!FeatureID) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(FeatureID==analyte) %>%
  ggplot(aes(x=StrainID, y=value, color=StrainID)) +
   stat_summary(geom="errorbar", width=0.2) +
   stat_summary(geom="point") +
  #geom_jitter(width=0.2, heigh=0) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~FeatureID, scales="free") +
   theme(legend.position = "none")
 print(p)
}
dev.off()

BA_wide %>%
  pivot_longer(!FeatureID) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(FeatureID %in% c("TCA/THCA","12-oxoDCA/7-oxoCDCA/6-oxoHDCA+5")) %>%
  mutate(FeatureID=factor(FeatureID, levels=c("TCA/THCA","12-oxoDCA/7-oxoCDCA/6-oxoHDCA+5"))) %>%
  ggplot(aes(x=StrainID, y=2^value)) +
  stat_summary(geom="errorbar", width=0.6) +
  stat_summary(geom="point") +
  facet_grid(FeatureID~., scales="free") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  xlab("") +
  ylab("Bile Acid Concentration (µM)")
ggsave("prelim_interest.pdf", height=3, width=4, useDingbats=F)



```

##heatmaps and bubble plots
```{r}

logBA_StBaseline <-
  BA_long %>%
  filter(Group == "Sterile") %>% 
  group_by(FeatureID, Group) %>%
  summarise(SterileLogConc=mean(log2Conc)) %>%
  arrange(desc(SterileLogConc))

BA_log2FC <-
  BA_long%>%
  group_by(FeatureID) %>% 
  left_join(logBA_StBaseline %>% dplyr::select(FeatureID, SterileLogConc)) %>%
  mutate(log2FC = log2Conc - SterileLogConc)


clust<-
  BA_log2FC %>%
  dplyr::select(SampleID, FeatureID, log2FC) %>%
  pivot_wider(names_from = FeatureID, values_from = log2FC) %>%
  column_to_rownames("SampleID") %>%
  t() %>%
  dist() %>%
  hclust(., method="average")

#log2fc heatmap of indiv strains only
BA_log2FC %>%
  filter(!grepl("Group", Group)) %>% 
  mutate(FeatureID=factor(FeatureID, levels = BAorderRev)) %>%
  #mutate(Group=factor(Group, levels=c("Sterile", "CA", "DCA", "CDCA", "LCA", "BApool", "EL_BApool"))) %>% 
  ggplot(aes(x=SampleID, y=FeatureID, fill=log2FC)) +
  geom_tile() +
  scale_fill_gradient2(low = "cornflowerblue", high="indianred", name="log2(Abundance)") +
  facet_wrap(nrow = 1,~Group, scales="free_x") +
  theme_q2r() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  xlab("") +
  ylab("")
ggsave("figures/BAheatmap_log2fc_indivstrains.pdf", height=10, width=25, useDingbats=F)


BA_log2FC %>%
  filter(!grepl("Group", Group)) %>% 
  mutate(FeatureID=factor(FeatureID, clust$labels[clust$order])) %>%
  ggplot(aes(x=SampleID, y=FeatureID)) +
  geom_point(aes(size = log2Conc, fill=log2FC), shape=21) +
  scale_fill_gradient2(low = "cornflowerblue", high="indianred", name="log2FC") +
  scale_size_continuous(range = c(0,6)) +
  ggtitle("log2FC vs sterile")+
  theme_q2r() +
  theme(axis.text.x = element_blank(), axis.ticks.x=element_blank()) +
  facet_wrap(nrow = 1,~Group, scales="free_x") +
  xlab("") +
  ylab("")
ggsave("figures/BAbubblemap_log2fc_indivstrains.pdf", height=10, width=25, useDingbats=F)

#log2fc heatmap of groups only
BA_log2FC %>%
  filter(grepl("Group", Group) | Group %in% c("Sterile", "Pool")) %>% 
  mutate(FeatureID=factor(FeatureID, clust$labels[clust$order])) %>%
  #mutate(Group=factor(Group, levels=c("Sterile", "CA", "DCA", "CDCA", "LCA", "BApool", "EL_BApool"))) %>% 
  ggplot(aes(x=SampleID, y=FeatureID, fill=log2FC)) +
  geom_tile() +
  scale_fill_gradient2(low = "cornflowerblue", high="indianred", name="log2(Abundance)") +
  facet_wrap(nrow = 1,~Group, scales="free_x") +
  theme_q2r() +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  xlab("") +
  ylab("")
ggsave("figures/BAheatmap_log2fc_groups.pdf", height=10, width=25, useDingbats=F)

BA_log2FC %>%
  filter(grepl("Group", Group) | Group %in% c("Sterile", "Pool")) %>% 
  mutate(FeatureID=factor(FeatureID, clust$labels[clust$order])) %>%
  ggplot(aes(x=SampleID, y=FeatureID)) +
  geom_point(aes(size = log2Conc, fill=log2FC), shape=21) +
  scale_fill_gradient2(low = "cornflowerblue", high="indianred", name="log2FC") +
  scale_size_continuous(range = c(0,6)) +
  ggtitle("log2FC vs sterile")+
  theme_q2r() +
  theme(axis.text.x = element_blank(), axis.ticks.x=element_blank()) +
  facet_wrap(nrow = 1,~Group, scales="free_x") +
  xlab("") +
  ylab("")
ggsave("figures/BAbubblemap_log2fc_groups.pdf", height=10, width=25, useDingbats=F)



```

##PCA
```{r}
BA_PCA<-
  BA_long %>%
  filter(!grepl("Group", Group)) %>% 
  dplyr::select(SampleID, FeatureID, log2Conc) %>%
  pivot_wider(names_from = "FeatureID", values_from = log2Conc) %>%
  as.data.frame() %>%
  column_to_rownames("SampleID") %>%
  prcomp()

loads<-
  BA_PCA$rotation %>% 
  as.data.frame() %>%
  rownames_to_column("FeatureID")%>%
  mutate(hyp=sqrt(PC1^2+PC2^2)) %>%
  top_n(15, hyp) %>%
  mutate(Group=NA)

scaler=20

BA_PCA$x %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  mutate(Group=gsub("_\\d+","", SampleID)) %>%
  ggplot(aes(x=PC1, y=PC2, fill=Group)) +
  geom_point(shape=21, size=3) +
  scale_fill_discrete()+
  geom_segment(data=loads, aes(x=0, xend=PC1*scaler, y=0, yend=PC2*scaler), arrow = arrow(length = unit(0.1,"cm"), type="closed")) +
  ggrepel::geom_text_repel(data=loads, aes(x=PC1*scaler, y=PC2*scaler, label=FeatureID)) +
  theme_q2r()
```

# Phylogeny tree
## make a tree
```{r}
straintree <- read.tree("/data/shared_resources/genomes/faas_phylophlan/RAxML_bestTree.faas_refined.tre")

sFMT_tree_k <- drop.tip(straintree, straintree$tip.label[!straintree$tip.label %in% BA_long$Group])

sFMT_tree_k %>% 
  ggtree::ggtree() + 
  ggtree::geom_tiplab()
```

## correlation between distance matrices (metabolite vs cophenetic)
```{r}
#pulling euclidean distance (cophenetic distance)

#correlation between distance matrices (metabolites vs phylogeny) procrust

#this analysis only contains indiv strain data

BA_PCA_dist<-
  BA_long %>%
  filter(grepl("JEB", Group)) %>% 
  group_by(FeatureID, Group) %>%
  summarise(MeanLogConc=mean(log2Conc)) %>%
  dplyr::select(Group, FeatureID, MeanLogConc) %>%
  pivot_wider(names_from = "FeatureID", values_from = MeanLogConc) %>%
  as.data.frame() %>%
  column_to_rownames("Group") %>%
  dist() %>% 
  as.matrix()


BA_cophenetic <-
  sFMT_tree_k %>% 
  cophenetic.phylo()

cor(as.vector(BA_PCA_dist), as.vector(BA_cophenetic))
#0.1818916

```

# BA labeling
```{r}

BA_long_labelled <-
  BA_long %>% 
  left_join(BAinfo %>% dplyr::select(FeatureID=QuanID, Labeling)) %>% 
  filter(Labeling == "Labeled")



BA_long_unlabelled <-
  BA_long %>% 
  left_join(BAinfo %>% dplyr::select(FeatureID=QuanID, Labeling)) %>% 
  filter(Labeling == "Unlabeled")

```


#Groups data
I identified 12-oxoDCA to be an interesting target because it is only produced by the pool but not by individual strains

##random forest to identify the important strains in production of 12-oxoDCA+5
```{r}
#read in the identifiers for the groups - strain members in each group
groups<-read_excel("/home/msk5737/BileAcids/221211_BA_sFMT_invitro/220111_BA_sFMT_invitro_samplemetadata.xlsx", sheet = 3)

#make a table of strain members in every group
grmem<-
  groups %>% 
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Group") %>%
  pivot_longer(!Group) %>%
  dplyr::select(Group, StrainID=value) %>%
  filter(!is.na(StrainID)) %>%
  bind_rows(
    tibble(Group=groups$Pool, StrainID=groups$Pool)
  )

#clean up table
grmem <-
  grmem %>% 
  mutate(jnk=1) %>% 
  pivot_wider(names_from = Group, values_from = "jnk", values_fill = 0) %>% bind_cols(., tibble(Sterile=0)) %>% 
  pivot_longer(!StrainID, names_to = "Group", values_to = "Present") %>%
  mutate(Group=factor(Group, levels=c("Sterile", grep("^JEB", unique(grmem$Group), value=TRUE), "Pool", grep("^Group", unique(grmem$Group), value=TRUE))))

#set prediction variable for random forest
RFpred <-
  grmem %>%
  filter(grepl("Group", Group)) %>%
  pivot_wider(names_from = "StrainID", values_from = "Present") %>%
  as.data.frame() %>%
  column_to_rownames("Group")

#set response variable for random forest
RFresp_12oxoDCA5 <-
BA_long %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%
  filter(FeatureID=="12-oxoDCA/7-oxoCDCA/6-oxoHDCA+5") %>%
  filter(grepl("Group", Group)) %>%
  group_by(Group) %>%
  summarize(Concentration_uM=mean(Concentration_uM))
  
RFpred_12oxoDCA5 <- 
  RFpred[RFresp_12oxoDCA5$Group,]

#run random forest
RFfit_12oxoDCA5 <- 
  randomForest::randomForest(RFpred_12oxoDCA5, RFresp_12oxoDCA5$Concentration_uM, importance=TRUE)

```

## make an upset plot
```{r}

#order the strains in the order of importance for presence of 12-oxoDCA+5
ImpOrder <- 
  RFfit_12oxoDCA5$importance %>%
  as.data.frame() %>%
  rownames_to_column("StrainID") %>%
  arrange(desc(`%IncMSE`))

#make the bottom portion of the upset plot
presence_hm<-
grmem %>%
  left_join(ImpOrder) %>% 
  mutate(ImpScore = case_when(Present == 0 ~ NA,
                              Present != 0 ~ `%IncMSE`)) %>% 
  ggplot(aes(x=Group, y=StrainID, fill=ImpScore)) +
  geom_tile() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  scale_fill_gradient(high= "black", low= "grey96", na.value = c("white")) +
  #theme(legend.position="none") +
  theme(axis.text.x=element_blank()) +
  theme(axis.ticks.x=element_blank())

#make the top portion of the upset plot (bar plot)
oxoDCA5_bar<-
BA_long %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%
  filter(FeatureID=="12-oxoDCA/7-oxoCDCA/6-oxoHDCA+5") %>%
  mutate(Group=factor(Group, levels=levels(grmem$Group))) %>%
  ggplot(aes(x=Group, y=Concentration_uM)) +
  stat_summary(geom="bar") +
  stat_summary(geom="errorbar", width=0, fun.data = mean_se) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  #theme(axis.text.x=element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  ylab("Concentration (µM)") +
  ggtitle("12-oxoDCA/7-oxoCDCA/6-oxoHDCA+5")

#stitch together and save as pdf
oxoDCA5_bar / presence_hm

ggsave("figures/12-oxoDCA5_upsetplot.pdf", height=10, width=10, useDingbats=F)

#look at unlabeled 12-oxoDCA
```


#SANDBOX
```{r, eval=FALSE, echo=FALSE}

RFresp_3oxoLCA5 <-
BA_long %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%
  filter(FeatureID=="3-oxoLCA+5") %>%
  filter(grepl("Group", Group)) %>%
  group_by(Group) %>%
  summarize(Concentration_uM=mean(Concentration_uM))
  
RFpred_3oxoLCA5 <- 
  RFpred[RFresp_3oxoLCA5$Group,]

RFfit_3oxoLCA5 <- 
  randomForest::randomForest(RFpred_3oxoLCA5, RFresp_3oxoLCA5$Concentration_uM, importance=TRUE)

RFfit_3oxoLCA5$importance %>%
  as.data.frame() %>%
  rownames_to_column("StrainID") %>%
  arrange(desc(`%IncMSE`))

BA_long %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%
  filter(FeatureID=="3-oxoLCA+5") %>%
  mutate(Group=factor(Group, levels=levels(grmem$Group))) %>%
  ggplot(aes(x=Group, y=Concentration_uM)) +
  stat_summary(geom="bar") +
  stat_summary(geom="errorbar", width=0, fun.data = mean_se) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  #theme(axis.text.x=element_blank()) +
  #theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  ylab("Concentration (µM)")

RFresp_12oxoDCA <-
BA_long %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%
  filter(FeatureID=="12-oxoDCA/7-oxoCDCA/6-oxoHDCA") %>%
  filter(grepl("Group", Group)) %>%
  group_by(Group) %>%
  summarize(Concentration_uM=mean(Concentration_uM))
  
RFpred_12oxoDCA <- 
  RFpred[RFresp_12oxoDCA$Group,]

RFfit_12oxoDCA <- 
  randomForest::randomForest(RFpred_12oxoDCA, RFresp_12oxoDCA$Concentration_uM, importance=TRUE)

RFfit_12oxoDCA$importance %>%
  as.data.frame() %>%
  rownames_to_column("StrainID") %>%
  arrange(desc(`%IncMSE`))

BA_long %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%
  filter(FeatureID=="CDCA+5") %>%
  mutate(Group=factor(Group, levels=levels(grmem$Group))) %>%
  ggplot(aes(x=Group, y=Concentration_uM)) +
  stat_summary(geom="bar") +
  stat_summary(geom="errorbar", width=0, fun.data = mean_se) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  #theme(axis.text.x=element_blank()) +
  #theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  ylab("Concentration (µM)")

BA_log2FC_sb <-
  BA_log2FC %>% 
  filter(FeatureID == "DCA")

tukey <-
  BA_long %>%
  filter(!grepl("Group", Group)) %>% 
  group_by(FeatureID) %>%
  filter(FeatureID == "DCA") %>% 
  do(
  aov(log2Conc~Group, data=.) %>%
  TukeyHSD() %>%
  broom::tidy()
  ) %>%
  filter(grepl("Sterile", contrast)) %>% 
  #filter(adj.p.value < 0.15) %>%
  dplyr::select(FeatureID, contrast, adj.p.value)

#make a change for git
```








#written by ben - save for reference
```{r, eval=FALSE, echo=FALSE}
# Filter Tree to contain tips from these experiments

straintree <- read.tree("/data/shared_resources/genomes/faas_phylophlan/RAxML_bestTree.faas_refined.tre")

filtered_tree_k <- drop.tip(straintree, straintree$tip.label[!straintree$tip.label %in% BA_long$Group])
```

```{r, eval=FALSE, echo=FALSE}
curves_k <- summary_curves_k %>%
  mutate(ID = paste0(StrainID, "-", Species))  %>%
  arrange(factor(StrainID, levels = filtered_tree_k$tip.label)) %>%
  filter(stat %in% c("art_k_fc","hcq_k_fc")) %>%
  dplyr::select(ID, stat, value) %>%
  pivot_wider(id_cols = ID, names_from = stat, values_from = value) %>%
  column_to_rownames("ID")
```

```{r, eval=FALSE, echo=FALSE}
## Assign New Tip Labels
new_tiplabels_k <- summary_curves_k %>%
  filter(StrainID %in% filtered_tree_k$tip.label) %>%
  mutate(ID = paste0(StrainID, "-", Species))  %>%
  arrange(factor(StrainID, levels = filtered_tree_k$tip.label)) %>%
  dplyr::select(ID, Species) %>%
  distinct() %>%
  pull(ID)
filtered_tree_k$tip.label <- new_tiplabels_k
```

```{r, eval=FALSE, echo=FALSE}
# Generate Tip Metadata
  mutate(ID = paste0(StrainID, "-", Species))  %>%
  filter(!StrainID %in% problem_taxa) %>%
  filter(ID %in% filtered_tree_k$tip.label) %>%
  dplyr::select(ID, Phylum, Species) %>%
  distinct() %>%
  arrange(factor(ID, levels = filtered_tree_k$tip.label)) %>%
  separate(Species, c("Genus","Spec"),remove = FALSE) %>%
  column_to_rownames("ID")
```

```{r, eval=FALSE, echo=FALSE}
# Tree with Fold Change and Circular Tiles
q2r_palette <- c("blue4", "olivedrab", "firebrick", "gold",
        "darkorchid", "steelblue2", "chartreuse1", "aquamarine")
p <- ggtree(filtered_tree_k, layout="circular")
p2 <- gheatmap(p, data = curves_k, offset = .25, width=0.2, low = "white", high = "darkred", color = "white", colnames = TRUE, font.size=3, colnames_position= "top", colnames_angle = 0, colnames_offset_y = 0, hjust = 0) + scale_fill_gradient2(na.value = "gray90", high = "darkred", low = "darkblue",name = "log2 Fold Change k (log2(Drug/Veh))") + xlim(NA,5)
p3 <- p2 + new_scale_fill()
p4 <- gheatmap(p3, tip_metadata %>% dplyr::select(Phylum) %>% filter(!Phylum %in% c("NA","Unknown")), offset=0, width=.1,
               colnames_angle=95, colnames_offset_y = .25) +
    scale_fill_manual(values = q2r_palette, na.value = "gray90", name = "Phylum")
p4
ggsave(paste0("Figures/heatmap_tree_k_fc_circ_tiles.pdf"), width=10, height=10, useDingbats=F)
```











#written by jordan - save for reference
```{r, eval=FALSE, echo=FALSE}

datmat<-
  mergeddata %>%
  filter(!grepl("Group", Group)) %>%
  dplyr::select(Analyte, SampleID, Concentration) %>%
  pivot_wider(names_from = SampleID, values_from = Concentration) %>%
  as.data.frame() %>%
  column_to_rownames("Analyte")

#QC Needed!


datmat<-datmat[rowSums(datmat)>0,]

datmat<-
datmat %>%
  rownames_to_column("Analyte") %>%
  pivot_longer(!Analyte, names_to = "SampleID", values_to="Concentration") %>%
  left_join(
        
        datmat %>%
          rownames_to_column("Analyte") %>%
          pivot_longer(!Analyte, names_to = "SampleID", values_to="Concentration") %>%
          group_by(Analyte) %>%
          filter(Concentration==min_nonzero(Concentration)) %>%
          ungroup() %>%
          dplyr::select(Analyte, MinConcentration=Concentration) %>%
          distinct()
  ) %>%
  mutate(log2Conc=log2(Concentration + (2/3*(MinConcentration)))) %>%
  dplyr::select(Analyte, SampleID, log2Conc) %>%
  pivot_wider(names_from = SampleID, values_from = log2Conc)

pc<-
datmat %>%
  column_to_rownames("Analyte") %>%
  t() %>%
  prcomp()

biplot(pc)


pc$x %>%
  as.data.frame() %>%
  rownames_to_column("StrainID") %>%
  separate(StrainID, c("StrainID", "Replicate"), sep="_") %>%
  ggplot(aes(x=PC1, y=PC2, color=StrainID)) +
  geom_point() +
  geom_line(aes(group=StrainID))



datmat %>%
  pivot_longer(!Analyte) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(Analyte=="TCA") %>%
  ggplot(aes(x=StrainID, y=value)) +
  geom_point() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~Analyte, scales="free")



datmat %>%
  pivot_longer(!Analyte) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(Analyte=="TDCA") %>%
  ggplot(aes(x=StrainID, y=value)) +
  geom_point() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~Analyte, scales="free")

pdf("skim.pdf", height=8, width=12, useDingbats=F)
for (analyte in datmat$Analyte){
 p<-
   datmat %>%
  pivot_longer(!Analyte) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(Analyte==analyte) %>%
  ggplot(aes(x=StrainID, y=value, color=StrainID)) +
   stat_summary(geom="errorbar", width=0.2) +
   stat_summary(geom="point") +
  #geom_jitter(width=0.2, heigh=0) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  facet_wrap(~Analyte, scales="free") +
   theme(legend.position = "none")
 print(p)
}
dev.off()

datmat %>%
  pivot_longer(!Analyte) %>%
  separate(name, c("StrainID", "Replicate"), sep="_") %>%
  filter(Analyte %in% c("TCA","12-oxoDCA+5")) %>%
  mutate(Analyte=factor(Analyte, levels=c("TCA","12-oxoDCA+5"))) %>%
  ggplot(aes(x=StrainID, y=2^value)) +
  stat_summary(geom="errorbar", width=0.6) +
  stat_summary(geom="point") +
  facet_grid(Analyte~., scales="free") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  xlab("") +
  ylab("Bile Acid Concentration (µM)")
ggsave("prelim_interest.pdf", height=3, width=4, useDingbats=F)

############################

dat<-read_csv("230127_BA_sFMT_invitro_ConcData.csv")
groups<-read_excel("/home/msk5737/BileAcids/221211_BA_sFMT_invitro/220111_BA_sFMT_invitro_samplemetadata.xlsx", sheet = 3)


hm<-
  groups %>% 
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Group") %>%
  pivot_longer(!Group) %>%
  dplyr::select(Group, StrainID=value) %>%
  filter(!is.na(StrainID)) %>%
  bind_rows(
    tibble(Group=groups$Pool, StrainID=groups$Pool)
  )


hm<-
hm %>% 
  mutate(jnk=1) %>% 
  pivot_wider(names_from = Group, values_from = "jnk", values_fill = 0) %>% bind_cols(., tibble(Sterile=0)) %>% 
  pivot_longer(!StrainID, names_to = "Group", values_to = "Present") %>%
  mutate(Group=factor(Group, levels=c("Sterile", grep("^JEB", unique(hm$Group), value=TRUE), "Pool", grep("^Group", unique(hm$Group), value=TRUE))))


pres<-
hm %>%
  ggplot(aes(x=Group, y=StrainID, fill=factor(Present))) +
  geom_tile() +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  scale_fill_manual(values=c("white","grey50")) +
  theme(legend.position="none") +
  theme(axis.text.x=element_blank()) +
  theme(axis.ticks.x=element_blank())

metab<-
dat %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%
  filter(FeatureID=="12-oxoDCA/7-oxoCDCA/6-oxoHDCA+5") %>%
  mutate(Group=factor(Group, levels=levels(hm$Group))) %>%
  ggplot(aes(x=Group, y=Concentration_uM)) +
  stat_summary(geom="bar") +
  stat_summary(geom="errorbar", width=0, fun.data = mean_se) +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  theme(axis.text.x=element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  ylab("Concentration (µM)")




metab / pres
ggsave("testplot.pdf", height=5, width=4)


preds<-
  hm %>%
  filter(grepl("Group", Group)) %>%
  pivot_wider(names_from = "StrainID", values_from = "Present") %>%
  as.data.frame() %>%
  column_to_rownames("Group")

response<-
dat %>%
  filter(!Group %in% c("JEB00356", "JEB00054")) %>%v
  filter(FeatureID=="12-oxoDCA/7-oxoCDCA/6-oxoHDCA+5") %>%
  filter(grepl("Group", Group)) %>%
  group_by(Group) %>%
  summarize(Concentration_uM=mean(Concentration_uM))
  
preds<-preds[response$Group,]


fit<-randomForest::randomForest(preds, response$Concentration_uM, importance=TRUE)

fit$importance %>%
  as.data.frame() %>%
  rownames_to_column("StrainID") %>%
  arrange(desc(`%IncMSE`))


```